Evolution of ORM Systems: Formal Foundations

### 2013-01-09 doc/description.tex
row and pair reformated

\paragraph{Property} represents a feature of  a class which is represented as a primitive type. The property can be mandatory, can have a default value and can represent a single value or a collection of values.
\begin{align*}
	\mathbf{Property} = <label, AppType, DefaultValue, Cardinality, Mandatory> \\ \\
	\mathbf{prop} : Label \times AppType \times String \times \mathbb{N^{*}} \times Boolean \rightarrow Property \\ \\
	\mathbf{name} : Property \rightarrow Label \\
	name(prop(l, t, val, n, m)) = l \\ \\
	\mathbf{type} : Property \times AppType \\
	type(prop(l, t, val, n, m)) = t \\ \\
	\mathbf{cardinality} : Property \rightarrow \mathbb{N^{*}} \\
	cardinality(prop(l, t, val, n, m)) = n \\ \\
	\mathbf{mandatory} : Property \rightarrow Boolean \\
	mandatory(prop(l, t, val, n, m)) = m  \\ \\
	l \in Label, t \in AppType, val \in String*, n \in \mathbb{N^{*}}, m \in Boolean 
\end{align*}

\paragraph {Association} represents a connection between two classes. It has a unique name and it contains a name of the class which is referenced by the association. The class which owns the association is consider to be the starting class of an association, referenced class is consider to be the ending class of an association. The cardinalities defines the multiplicity of the association ends.
\begin{align*}
	\mathbf{Association} = <label, classRef, startCardinality, endCardinality> \\ \\
	\mathbf{assoc} : Label \times Label \times \mathbb{N} \times \mathbb{N} \rightarrow Association\\
	\mathbf{name} : Association \rightarrow Label \\
	name(assoc(l, ref, n_1, n_2)) = l\\ \\
	\mathbf{reference} : Association \rightarrow Label \\
	reference(assoc(l, ref, n_1, n_2)) = ref\\ \\
	\mathbf{startCardinality} : Association \rightarrow \mathbb{N} \\
	startCardinality(assoc(l, ref, n_1, n_2)) = n_1\\ \\
	\mathbf{endCardinality} : Association \rightarrow \mathbb{N} \\
	endCardinality(assoc(l, ref, n_1, n_2)) = n_2 \\ \\
l, ref \in Label,  n_1, n_2 \in \mathbb{N}
\end{align*}

@@

\paragraph{Column} columns define possible data values and types which can be part of a table record.
\begin{align*}
	\mathbf{Column} = <label, DbType, DefaultValue, Constraint*> \\ \\
	\mathbf{col} : Label \times DbType \times String \times Constraints \rightarrow Column \\ \\
	\mathbf{name} : Column \rightarrow Label  \\
	name(col(l, t, val, cons)) = l  \\ \\
	\mathbf{type} : Column \rightarrow DbType  \\
	type(col(l, t, val, cons)) = t  \\ \\
	\mathbf{constraints} : Column \rightarrow Constraints  \\
	constraints(col(l, t, val, cons)) = cons  \\ \\
l \in Label, t = DbType, val = Strings, cons = Constraints
\end{align*}

\paragraph{Foreign key}
\begin{align*}
	\mathbf{ForeignKey} = <label, tableRef, Constraint*> \\ \\
	\mathbf{fk} : Label \times Label \times Constraints \rightarrow ForeignKey \\ \\
	\mathbf{name} : ForeignKey \rightarrow Label \\
	name(fk(l, tRef, cons)) = lab  \\ \\
	\mathbf{reference} : ForeignKey \rightarrow Label  \\
	reference(fk(l, tRef, cons)) = tRef  \\ \\
	\mathbf{constaints} : ForeignKey \rightarrow Constraint*  \\
	constaints(fk(l, tRef, cons)) = cons  \\ \\
l, tRef \in Label, cons \in Constraints
\end{align*}

@@

\subsubsection{Model of Stored Data}
A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of reference to a concrete table and a set of value pairs, which represents concrete values of concrete columns.
\begin{align*}
	\mathbf{Row} = < refTable, pair* > \\
	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
	\mathbf{refTable} : Row \rightarrow Label \\
	refTable(row(l, pairs) = l \\ \\
	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
 l \in Label, ps = Pair*, t \in tables(d), d \in Database
\end{align*}
There functions $contains$ verify if the table contains data in context of concrete database.

\begin{align*}
	\mathbf{Pair} = < refCol, value > \\
	[Label \times Value] \rightarrow Pair\\ \\
	\mathbf{reference} : Pair \rightarrow Label \\
	reference([l,val]) = l \\ \\
	\mathbf{value} : Pair \rightarrow Label \\ 
	value([l,val]) = val \\ \\
	\mathbf{pairByRef} : Label \times Row \rightarrow Pair \\
	pairByRef(l,i) = \begin{cases}
			\inference{\exists p \in pairs(i) : reference(p) = l}{p} \\ \\
			\varnothing
		\end{cases}\\
	l \in Label, val \in Value
\end{align*}
The $refCol$ is a name of a column, a foreign key or its primary key and $value$ represents concrete value stored in the database. Each row in a database has to refer to a table in database schema and has a structure corresponding to the one defined by a table, otherwise the database is inconsistent. The consistency of database is then:
\begin{align*}
	\mathbf{consistent} : Database \rightarrow Boolean \\
	consistent(d) = \begin{cases}
 		\inference{\forall r \in data(d), \exists t \in tables(d) : reference(r) = name(t) \\ \wedge \forall p \in pairs(r) :\\ \exists u \in columns(t) \cup foreignKeys(t) : \\ reference(p) = name(u)}{true}
 	\\ \\
 	false
 \end{cases}
\end{align*}

\subsection{Extension of Software Definition}

\begin{align*}
 owningClass : Property \times Application \rightarrow Class  \\ owningClass(p, a) = c, c \in classes(a) \wedge p \in properties(c) \\
  owningClass : Association \times Application \rightarrow Class  \\
owningClass(as, a) = c, c \in classes(a) \wedge as \in associations(c) \\ \\
owningTable : Column \times Database \rightarrow Table  \\
owningTable(c, d) = tab, tab \in tables(d) \wedge c \in columns(d) \\ \\
owningTable : ForeignKey \times Database \rightarrow Table  \\
owningTable(fk, d) = tab, tab \in tables(d) \wedge fk \in foreignKeys(d)\\ \\ 
\end{align*}

The database stores instances of the application, thus the $instantiated$ function has to be defined:

@@

\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Column* \\ \\
	\rho_{sp}(p) = \begin{cases}
 		\inference{cardinality(p) = 1 \wedge !\;mandatory(p)
 			\\ \wedge \exists t : t = table(name(owningClass(p)), d) 
		 }{\begin{gathered}
	  		column(name(p), \rho_t(type(p)), \varnothing )
		\end{gathered}} \\ \\
  		\inference{cardinality(p) = 1 \wedge mandatory(p)
			 \\ \wedge \exists t : t = table(name(owningClass(p)), d)
			\\ \wedge \not instanciated(owningClass(p))
		}{\begin{gathered}
 	 		column(name(p), \rho_t(type(p)), NOTNULL ) \\ 
		\end{gathered}}
 	\end{cases}
\end{align*}
\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Table* \\ \\
	\rho_{cp}(p) = \begin{cases}
 	\inference{cardinality(p) > 1 \wedge !\;mandatory(p) \\ 
 		\wedge 	\exists t : t = table(name(owningClass(p)), d)
 		}{\begin{gathered}
 			table(name(p), primaryKey("Id"), \\ column("value", 			\rho_t(type(p)), \varnothing), \\ foreignKey(name(p), 			name(owningClass(p)), NOTNULL) \\
	\end{gathered}}\\ \\
 	\inference{ cardinality(p) > 1 \wedge mandatory(p) \\ 
 		\wedge \exists t : t = table(name(owningClass(p)), d) 
 		}{\begin{gathered}  
 			table(name(p), primaryKey("Id"), \\ 
 			column("value", \rho_t(type(p)), NOTNULL), \\ 			foreignKey(name(p), name(owningClass(p)), NOTNULL)
	 \end{gathered}}
 	\end{cases}
\end{align*}

>>>>

\paragraph{Property} represents a feature of  a class which is represented as a primitive type. The property can be mandatory, can have a default value and can represent a single value or a collection of values.
\begin{align*}
	\mathbf{Property} = <label, AppType, DefaultValue, Cardinality, Mandatory> \\ \\
	\mathbf{prop} : Label \times AppType \times String \times \mathbb{N^{*}} \times Boolean \rightarrow Property \\ \\
	\mathbf{name} : Property \rightarrow Label \\
	name(prop(l, t, val, n, m)) = l \\ \\
	\mathbf{type} : Property \times AppType \\
	type(prop(l, t, val, n, m)) = t \\ \\
	\mathbf{cardinality} : Property \rightarrow \mathbb{N^{*}} \\
	cardinality(prop(l, t, val, n, m)) = n \\ \\
	\mathbf{mandatory} : Property \rightarrow Boolean \\
	mandatory(prop(l, t, val, n, m)) = m  \\ \\
	\mathbf{owningClass} : Property \times Application \rightarrow Class  \\ 	
	owningClass(p, a) = \inference{\exists c \in classes(a) \wedge p \in properties(c)}{c}  \\ \\
	l \in Label, t \in AppType, val \in String*, n \in \mathbb{N^{*}}, m \in Boolean 
\end{align*}

\paragraph {Association} represents a connection between two classes. It has a unique name and it contains a name of the class which is referenced by the association. The class which owns the association is consider to be the starting class of an association, referenced class is consider to be the ending class of an association. The cardinalities defines the multiplicity of the association ends.
\begin{align*}
	\mathbf{Association} = <label, classRef, startCardinality, endCardinality> \\ \\
	\mathbf{assoc} : Label \times Label \times \mathbb{N} \times \mathbb{N} \rightarrow Association\\
	\mathbf{name} : Association \rightarrow Label \\
	name(assoc(l, ref, n_1, n_2)) = l\\ \\
	\mathbf{reference} : Association \rightarrow Label \\
	reference(assoc(l, ref, n_1, n_2)) = ref\\ \\
	\mathbf{startCardinality} : Association \rightarrow \mathbb{N} \\
	startCardinality(assoc(l, ref, n_1, n_2)) = n_1\\ \\
	\mathbf{endCardinality} : Association \rightarrow \mathbb{N} \\
	endCardinality(assoc(l, ref, n_1, n_2)) = n_2 \\ \\
	\mathbf{owningClass} : Association \times Application \rightarrow Class  \\
	owningClass(as, a) = \inference{\exists c \in classes(a) \wedge as \in associations(c)}{c}  \\ \\	
	l, ref \in Label,  n_1, n_2 \in \mathbb{N}, as = Association
\end{align*}

@@

\paragraph{Column} columns define possible data values and types which can be part of a table record.
\begin{align*}
	\mathbf{Column} = <label, DbType, DefaultValue, Constraint*> \\ \\
	\mathbf{col} : Label \times DbType \times String \times Constraints \rightarrow Column \\ \\
	\mathbf{name} : Column \rightarrow Label  \\
	name(col(l, t, val, cons)) = l  \\ \\
	\mathbf{type} : Column \rightarrow DbType  \\
	type(col(l, t, val, cons)) = t  \\ \\
	\mathbf{constraints} : Column \rightarrow Constraints  \\
	constraints(col(l, t, val, cons)) = cons  \\ \\
	\mathbf{owningTable} : Column \times Database \rightarrow Table  \\
	owningTable(c, d) = \inference{\exists tab \in tables(d) \wedge c \in columns(tab)}{tab} \\ \\
l \in Label, t = DbType, val = String,\\ cons = Constraints, c = Column, d = Database
\end{align*}

\paragraph{Foreign key}
\begin{align*}
	\mathbf{ForeignKey} = <label, tableRef, Constraint*> \\ \\
	\mathbf{fk} : Label \times Label \times Constraints \rightarrow ForeignKey \\ \\
	\mathbf{name} : ForeignKey \rightarrow Label \\
	name(fk(l, tRef, cons)) = lab  \\ \\
	\mathbf{reference} : ForeignKey \rightarrow Label  \\
	reference(fk(l, tRef, cons)) = tRef  \\ \\
	\mathbf{constaints} : ForeignKey \rightarrow Constraint*  \\
	constaints(fk(l, tRef, cons)) = cons  \\ \\
	\mathbf{owningTable} : ForeignKey \times Database \rightarrow Table  \\
	owningTable(fk, d) = \inference{\exists tab \in tables(d) \wedge fk \in foreignKeys(tab)}{tab}\\\\
	l, tRef \in Label, cons = Constraint*, d = Database, fk = ForeignKey
\end{align*}

@@

\subsubsection{Model of Stored Data}
A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of reference to a concrete table and a set of value pairs, which represents concrete values of concrete columns.
\begin{align*}
	\mathbf{Row} = < refTable, pair* > \\
	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
	\mathbf{refTable} : Row \rightarrow Label \\
	refTable(row(l, ps) = l \\ \\
	\mathbf{pairs} : Row \rightarrow Pair* \\
	pairs(row(l, ps)) = ps \\ \\
	\mathbf{pairByRef} : Label \times Row \rightarrow Pair \\
	pairByRef(l,i) = \begin{cases}
		\inference{\exists p \in pairs(i) : reference(p) = l}{p} \\ \\
			\varnothing
		\end{cases}\\
	 l \in Label, ps = Pair*, i = Row
\end{align*}
The value stored in concrete column of the row is represented by pair, where the $refCol$ is a name of the column, a foreign key or its primary key and $value$ represents concrete value .
\begin{align*}
	\mathbf{Pair} = < refCol, value > \\
	[Label \times Value] \rightarrow Pair\\ \\
	\mathbf{reference} : Pair \rightarrow Label \\
	reference([l,val]) = l \\ \\
	\mathbf{value} : Pair \rightarrow Label \\ 
	value([l,val]) = val \\ \\
	l \in Label, val \in Value
\end{align*}
There functions $contains$ verify if the table contains data in context of concrete database.
\begin{align*}
	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
 l \in Label, ps = Pair*, t \in tables(d), d \in Database
\end{align*}
Each row in a database has to refer to a table in database schema and has a structure corresponding to the one defined by a table, otherwise the database is inconsistent. The consistency of database is then:
\begin{align*}
	\mathbf{consistent} : Database \rightarrow Boolean \\
	consistent(d) = \begin{cases}
 		\inference{\forall r \in data(d), \exists t \in tables(d) : reference(r) = name(t) \\ \wedge \forall p \in pairs(r) :\\ \exists u \in columns(t) \cup foreignKeys(t) : \\ reference(p) = name(u)}{true}
 	\\ \\
 	false
 \end{cases}
\end{align*}

\subsection{Extension of Software Definition}
The database stores instances of the application, thus the $instantiated$ function has to be defined:

@@

\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Column* \\ \\
	\rho_{sp}(p) = \begin{cases}
		\inference{p = \varnothing}{\varnothing}\\
		\inference{|p| > 1}{\rho_{sp}(head(p)) \cup \rho_{sp}(tail(p))}\\\\
 		\inference{cardinality(p) = 1 \wedge !\;mandatory(p)
 			\\ \wedge \exists t : t = table(name(owningClass(p)), d) 
		 }{\begin{gathered}
	  		column(name(p), \rho_t(type(p)), \varnothing )
		\end{gathered}} \\ \\
  		\inference{cardinality(p) = 1 \wedge mandatory(p)
			 \\ \wedge \exists t : t = table(name(owningClass(p)), d)
			\\ \wedge \not instanciated(owningClass(p))
		}{\begin{gathered}
 	 		column(name(p), \rho_t(type(p)), NOTNULL ) \\ 
		\end{gathered}}
 	\end{cases}
\end{align*}
\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Table* \\ \\
	\rho_{cp}(p) = \begin{cases}
 	\inference{cardinality(p) > 1 \wedge !\;mandatory(p) \\ 
 		\wedge 	\exists t : t = table(name(owningClass(p)), d)
 		}{\begin{gathered}
 			table(name(p), primaryKey("Id"), \\ column("value", 			\rho_t(type(p)), \varnothing), \\ foreignKey(name(p), 			name(owningClass(p)), NOTNULL) \\
	\end{gathered}}\\ \\
 	\inference{ cardinality(p) > 1 \wedge mandatory(p) \\ 
 		\wedge \exists t : t = table(name(owningClass(p)), d) 
 		}{\begin{gathered}  
 			table(name(p), primaryKey("Id"), \\ 
 			column("value", \rho_t(type(p)), NOTNULL), \\ 			foreignKey(name(p), name(owningClass(p)), NOTNULL)
	 \end{gathered}}
 	\end{cases}
\end{align*}
